---
title: "Lab 3: COVID-19"
subtitle: "ESS 330"
author:
  - name: "https://github.com/bellaconrad/"
  - email: "bellaco@colostate.edu"
format: 
 html:
   self-contained: true
execute: 
  echo: true
---
```{r}
library(zoo)
library(flextable)
library(tidyverse)
library(dplyr)
library(tidyr)
library(skimr)
```

# Question 1
```{r}
library(tidyverse)
url = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
covid_data = read_csv(url)
```

# Question 2
```{r}
txt <- "2022-02-01"
class(txt)
date_example <- as.Date(txt)
class(date_example)
```

```{r}
my.date <- as.Date("2022-02-01")
my.state <- "Colorado"
```

```{r}
colorado <- covid_data %>% 
  filter(state == my.state) %>% 
  group_by(county) %>% 
  arrange(date) %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths)) %>% 
  ungroup()

```

```{r}
filter(colorado, date == my.date) %>% 
  slice_max(cases, n = 5) %>% 
  select(Date = date, County = county, Cases = cases) %>% 
  flextable() %>% 
  set_caption(caption = "Most Total Cases")

filter(colorado, date == my.date) %>% 
  slice_max(cases, n = 5) %>% 
  select(Date = date, County = county, Cases = new_cases) %>%
  flextable() %>% 
  set_caption(caption = "Most New Cases")
```

# Question 3
```{r}
pop_url <- 'https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv'
pop_data <- read_csv(pop_url) %>% 
  filter(COUNTY != "000") %>% 
  mutate(fips = sprintf("%05s", paste0(STATE, COUNTY))) %>% 
  select(fips, contains("NAME"), contains("2021"))
```

```{r}
names(pop_data)
dim(pop_data)
nrow(pop_data)
str(pop_data)
glimpse(pop_data)
skim(pop_data)
# There are 19 columns and 3144 rows. 3 of the columns are characters and 16 are numeric. The death column matches the COVID death column, but this one is specifically for the year 2021.
```

```{r}
co_pop <- pop_data %>% 
  filter(STNAME == "Colorado") %>% 
  group_by(CTYNAME)

pop_range <- range(co_pop$`POPESTIMATE2021`, na.rm = TRUE)

cat("Range of populations in Colorado counties in 2021:", pop_range)
```

```{r}
pop_and_covid <- colorado %>% 
  left_join(pop_data, by = "fips") %>% 
  mutate(
    cumulative_cases = ifelse(!is.na(cases) & !is.na(POPESTIMATE2021), cases / POPESTIMATE2021, NA),
    new_cases = ifelse(!is.na(new_cases) & !is.na(POPESTIMATE2021), new_cases / POPESTIMATE2021, NA),
    new_deaths = ifelse(!is.na(new_deaths) & !is.na(POPESTIMATE2021), new_deaths / POPESTIMATE2021, NA)
  )
```

```{r}
most_cum_cases <- pop_and_covid %>% 
  filter(date == my.date) %>% 
  slice_max(cumulative_cases, n = 5) %>% 
  select(date, county, cases, deaths, cumulative_cases)

flextable(most_cum_cases) %>% 
  set_header_labels(
    cumulative_cases = "Cumulative Cases per capita") %>% 
  set_caption("Top 5 Colorado Counties with Most Cumulative Cases Per Capita ") %>% 
  print()
```


```{r}
most_new_cases <- pop_and_covid %>% 
  filter(date == my.date) %>% 
  slice_max(new_cases, n = 5) %>% 
  select(date, county, cases, deaths, new_cases)

flextable(most_new_cases) %>% 
  set_header_labels(
    new_cases = "New Cases per capita") %>% 
  set_caption("Top 5 Colorado Counties with Most New Cases Per Capita ") %>% 
  print()
```

# Question 4

```{r}
latest_date <- max(pop_and_covid$date, na.rm = TRUE)

pop_and_covid %>%
  filter(date >= (latest_date - 13)) 

last_14_days <- pop_and_covid %>% 
  group_by(county) %>% 
  summarize(
    total_new_cases_14d = sum(new_cases, na.rm = TRUE),
    population = first(POPESTIMATE2021)
    ) %>%
  mutate(cases_per_100k = (total_new_cases_14d/population) *100000)

watchlist_count <- last_14_days %>% 
  filter(cases_per_100k>100) %>% 
  nrow()

top_5_counties <- last_14_days %>% 
  slice_max(order_by = cases_per_100k, n = 5)

print(last_14_days)
print(paste("Counties with cases per 100k > 100:", watchlist_count))
print(top_5_counties)

```

# Question 5
```{r}
CO_death_analysis <- pop_and_covid %>% 
  filter(state == "Colorado", year(date) == 2021) %>% 
  group_by(county) %>% 
  summarize(
    DEATHS2021 = first(DEATHS2021),
    deaths = first(deaths)) %>% 
  mutate(percentcoviddeath = (deaths/DEATHS2021)*100) %>% 
  mutate(high_impact_counties = (percentcoviddeath >= 20))

plot1 <- ggplot(CO_death_analysis, aes(x = county, y = percentcoviddeath)) +
  geom_col(fill = "darkred") +
  labs(title = "Colorado Counties where over 20% of Deaths were Caused by COVID-19 in 2021",
       x = "County",
       y = "Percentage of Total Deaths Caused by COVID-19") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
ggsave("Colorado_Counties_20_COVID_deaths.png", plot = plot1, width = 10, height = 6, dpi = 300)
print(plot1)
```


# Question 6
# Scaling by population can make certain states with larger cities look worse. For example, New York had very strict COVID policies in place, but since New York City is one of the largest cities in the US, it most likely caused there to also be a higher number of COVID cases.


```{r}

state_covid = covid_data %>% 
  group_by(date, state) %>% 
  summarise(cases = sum(cases)) %>% 
  filter(state %in% c('New York', "Ohio", 'Colorado', "Alabama")) %>% 
  group_by(state) %>% 
  mutate(newCases = cases - lag(cases),
         roll = zoo::rollmean(newCases, k = 7, align = "right", fill = NA)) %>% 
  ungroup()

roll_left <- rollmean(state_covid$newCases, k = 3, fill = NA, align = "left")

roll_right <- rollmean(state_covid$newCases, k = 3, fill = NA, align = "right")

plot2 <- ggplot(state_covid, aes(x = date)) +
  geom_col(aes(y = newCases), fill = "pink", col = NA) +
  geom_line(aes(y = roll), col = "darkred", linewidth = 1) +
  theme_linedraw()+
  facet_wrap(~state, nrow = 2, scales = "free_y") +
  labs(title = "Cumulative COVID-19 Cases",
       x = "Date", y = "Case Count")
ggsave("Cumulative_COVID_cases.png", plot = plot2, width = 10, height = 6, dpi = 300)
print(plot2)

```

```{r}
pp = pop_data %>% 
  group_by(STNAME) %>% 
  summarise(state_pop = sum(POPESTIMATE2021)) %>% 
  inner_join(state_covid, by = c("STNAME"="state")) %>% 
  mutate(perCap = newCases / state_pop) %>% 
  group_by(STNAME) %>% 
  mutate(roll = zoo::rollmean(perCap, k = 7, align = "right", fill = NA)) %>% 
  ungroup()
```

```{r}
plot3 <- ggplot(pp, aes(x = date)) +
  geom_line(aes(y = roll, col = STNAME), size = 1) +
  theme_linedraw() +
  labs(title = "Cumulative COVID-19 Cases",
       x = "Date", y = "Case Count")
ggsave("Cumulative_COVID_cases2.png", plot = plot3, width = 10, height = 6, dpi = 300)
print(plot3)
```


# Question 7
# The movement of the COVID-19 weighted mean throughout the USA I think is due to the fact that the midwest had the least amount of policies regarding COVID protections, like wearing masks or getting vaccinated, so it makes sense that the mean would be concentrated there. Some possible drivers include policy and policy enforcement, political attitudes, how much these cities/ states shut down compared to others, and much more.


```{r}
meta = read_csv('https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/county-centroids.csv') %>% 
  inner_join(covid_data) %>% 
  group_by(date) %>% 
  summarise(wmX_c = sum(LON*cases) / sum(cases),
            wmY_c = sum(LAT*cases) / sum(cases),
            cases = sum(cases)) %>% 
  arrange(date) %>% 
  mutate(d = 1:n())

plot4 <- ggplot(meta) +
  borders("state",fill = "gray90", colour = "white") +
  geom_point(aes(x = wmX_c, y = wmY_c, size = cases), color = "red", aplpha =.25) +
  theme_linedraw() +
  labs(color = "Time",
       size = "Cases",
       x = "", y = "",
       title = "Weighted Center of COVID-19 Cases") +
  theme(legend.position = "none")

borders("state", fill = "gray90", colour = "white")

ggsave("Weighted_US_Center_COVID.png", plot = plot4, width = 10, height = 6, dpi = 300)
print(plot4)
```

